---
title: "Notes Practice"
author: "Grace"
date: "2024-09-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Practicals I

**Note some shortcuts**\
- Code chunk: Ctrl + Alt + i

```{r packages}
install.packages("ade4")
library("ade4")
```

## Basic functions

```{r as a calculator}
pi # pi value
trunc(pi) # return the truncating value
ceiling(pi) # return the smallest integers not less than the numeric argument
floor (pi)  # return the largest integers not greater than the numeric argument 
round(pi,2) # round the value in its first argument to the specified number of decimal places
```

```{r functions}
c(1,2,3,4,5) # vector containing 5 integers
Temp <- c(15,18,23,19,24)
Temp
Far <- (Temp*(9/5))+32
mean(Far)
```

```{r functions}
c(1,2,3,4,5) # vector containing 5 integers
Temp <- c(15,18,23,19,24)
Temp
Far <- (Temp*(9/5))+32
mean(Far)
```

```{r regular sequences}
4:19 # generate a vector containing the sequence of all integers between

seq(from=0, to=12, by=2) # regular sequence with beginning, end, step size

rep(c(12,35,48),each=2,time = 2) # repeated sequence, specify the number of times each element is repeated (argument each) and the number of times the sequence is repeated (argument times). 

```

### Matrix

**matrix()** creates an array of data from a set of values or a vector with a given number of columns (argument ncol) and/or a given number of rows (argument nrow), and the order of filling (argument byrow; by column if byrow = FALSE and by line if byrow = TRUE).

```{r matrix}
mat<-matrix(1:9, ncol=3, byrow=FALSE)
mat

#set column and row names
colnames(mat)<- c("a","b","c")
rownames(mat)<- c("q","r","s")
mat
```

```{r extracting values}
mat[1,2] #extract values

#extract rows or columns
mat[1,]
mat[,1]

#extract the 2nd and 3rd row and the 1st and 2nd column
mat[c(2:3),c(1:2)]

```

```{r manipulating values}
#replace a value of the matrix
mat[1,2]<- 35; mat

# remove a value, a row, a column or a subsection of the array
mat2 <- mat[-1,]; mat2

#make an operation on a row, a column or a subsection of the matrix 
#calculate the mean of the 3rd column without considering the value of the 2nd row
mean(mat[-2,3])
```

### Logical operators

> , \>=, \<, \<=, ==, and !=

```{r extracting values with logical operators}

# extract (1) all values lesser than or equal to 3 and (2) all values strictly lesser than 9 and greater than or equal to 7
mat[mat<=3]
mat[mat<9&mat>=7]
```

```{r replacing values with logical operators}
#replace by 50 the values lower than or equal to 3
mat[mat<=3] <- 50; mat

```

### Data frame

```{r matrix of numeric and categorical data}
data = data.frame(letter=c("a","b","c","d","e","f"), order = seq(from=2, to=12, by=2), type = c("fries","burger","fish", "squid", "kebab", "spaghetti"))
data
```

## Getting help with R

```{r}
help.search("Fisher") 
help.start()

#search with keywords
?fisher.test()
```

## Import and manipulate a data spreadsheet

```{r import}
irisdata<- read.table("irisdata.txt", header=TRUE,row.names=1,dec=",")
```

```{r basic functions}
names(irisdata) #columns
head(irisdata) #first 6 lines
dim(irisdata) #number of line, column number
```

```{r accessing values}
irisdata[,3] #3rd column

#or using the syntax dataframe_name$variable_name
irisdata$Petal.Length
```

### Attach function

The database is attached to the R search path. This means that the database is searched by R when evaluating a variable, so objects in the database can be accessed by simply giving their names.

To detach your object (before setting another) use the function **detach ()**. If 2 files have the same variable names, only variables of the last file will be stored in memory.

```{r attach() }
attach(irisdata)
Petal.Length
Sepal.Width
```

### Manipulating Data

```{r change name of column or row}
colnames(irisdata)<-c("Lsepal", "Wsepal", "Lpetal", "Wpetal", "species")
head(irisdata)
```

```{r remove all row with missing values}
irisdataclean <- na.omit(irisdata)
dim(irisdataclean)
attach(irisdataclean)
```

```{r select observations following criteria}
#samples with petal length greater than or equal to 6.5
#you want to get the samples/rows that is why it is [n,]
#cond is just a variable name
cond.Lpetal <- irisdataclean[Lpetal>=6.5,]
cond.Lpetal
```

```{r select observations following criteria}
#select iris samples with petal length greater than or equal to 6.5 and sepal width lesser than 3  
#don't forget the ,
irisdataclean[Lpetal>=6.5&Wsepal<3,]
```

```{r subsection of an array}
#create a table containing only petal and sepal length of Iris setosa
setosa <- irisdataclean[species=="setosa",c("Lpetal", "Lsepal")]
setosa
```

### Ordering

```{r order a data table}
# order() - ascending order 
irisdataclean[order(Lpetal),]
# rev() - descending order
irisdataclean[rev(order(Lpetal)),]
```

### Combining

```{r combine different data frame}
#rbind() - combine 2 or more tables by rows
irisdataclean[species== "setosa",c(1,2,5)]->setosa 
irisdataclean[species== "virginica",c(1,2,5)]->virginica 
rbind(setosa,virginica) 

#bind() - combine 2 or more tables by columns
irisdata[species=="setosa",c(1,2)]->sepal 
irisdata[species=="setosa",c(3,4)]->petal 
cbind(sepal,petal) 
```

### Basic Statistics

**summary()** provides descriptive statistics (min, 1st quartile, median, 3rd quartile, max) for quantitaive variables and the count of each category for qualitative variables.

```{r summary}

#correct the data types first
irisdataclean$Lsepal <- as.numeric(irisdataclean$Lsepal)
irisdataclean$Wsepal <- as.numeric(irisdataclean$Wsepal)
irisdataclean$Lpetal <- as.numeric(irisdataclean$Lpetal)
irisdataclean$Wpetal <- as.numeric(irisdataclean$Wpetal)
irisdataclean$species <- as.factor(irisdataclean$species)

summary(irisdataclean)
```

```{r other basic stat}
quantile(irisdataclean[,"Lpetal"],0.25)
median(irisdataclean[,"Lpetal"])
sd(irisdataclean[,"Lpetal"])
```

**tapply()** is used to apply an operation (mean, sd…) on a quantitative variable by category of a qualitative variable.

Usage: **tapply (X, INDEX, FUN)**, with X: the quantitative variable; INDEX: the qualitative variable; FUN : the name of the function to make the operation.

```{r example}
# calculate for each species the mean of the sepal length
tapply(X=Lsepal,INDEX=species, FUN=mean)

# calculate for each species the mean of the petal length
tapply(Lpetal, species, mean)

```

## Plots

```{r plot function}
plot(Lpetal)
```

argument **main = "text"** is used to add a title\
symbol **\~** means “function of”.

```{r plot function}
plot(Lpetal~Lsepal,main="Petal Length") 
```

### Usage of the plot() function

plot ( X \~ Y, type = " type of plot "\* , main = " overall title " , sub = " sub title ", xlab = " x axis title ", ylab = " y axis title " , xlim = c (min , max ) , ylim = c (min , max ) )

-   Possible types : " p " for points, " l " for lines, " b " for both, " s " for stair steps...

```{r axis}
plot(Lpetal,axes=FALSE)
# axes=FALSE means that no axes will be drawn by default
axis(2) # Add y-axis, 1 is the x axis

plot(Lpetal, axes=FALSE, xlab="observations", ylab="petal length", ylim=c(0,10))
axis(1,col="grey",font.axis=2)
axis(2,col="grey",las=2)

```

```{r abline}
plot(Lpetal~Lsepal)
abline(lm(Lpetal~Lsepal,irisdataclean),lwd=3)
#lwd is the line width
# abline with equation y = b*x+a
abline(a=0, b=1, col="red", lwd=2)

#regression only on the species setosa
setosadata <- irisdataclean[species=="setosa",]
plot(Lpetal~Lsepal, setosadata)
abline(lm(Lpetal~Lsepal, setosadata)) + title(main="Iris setosa")
```

```{r points}
# make a graph for Iris setosa, then add the 2 other species with different colors
plot(Lpetal~Lsepal,setosadata, xlim=c(4,8), ylim=c(0,7))
points(Lpetal~Lsepal,irisdataclean[species=="virginica",], col="red")
points(Lpetal~Lsepal, irisdataclean[species=="versicolor",], col="green")

```
```{r lines}
setosaorder<-setosadata[order(setosadata$Lpetal),]
attach(setosaorder) 
plot(Lpetal)
lines(Lpetal,lwd=2) 
```
```{r hist}
hist(setosadata$Lpetal,breaks=5)
```

```{r hist}
hist(setosadata$Lpetal,breaks=c(1,1.2,1.4,1.6,1.8,2))

#remove y axis, add title
hist(setosadata$Lpetal,breaks=5,axes=FALSE,main="petal length",col="grey")
axis(1,col.axis="grey") 

```

```{r boxplot}
boxplot(irisdataclean$Lpetal, main="Petal Length")

#with different variables
boxplot(Lpetal, Wpetal, Lsepal, Wsepal, main="Iris boxplot", names=c("length petal", "width petal", "length sepal", "widthsepal"))

#by species
boxplot(irisdataclean$Lpetal~irisdataclean$species, main="petal length", names=c("I. setosa","I. diversicolor", "I. virginica"))
```

