---
title: Practical Session no. 2
description: training in the manipulation of R
required_packages: ggplot2, hrbrthemes
author: Grace Catapang
date: 2024-09-19
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Practical Session 2

```{r packages}
library("ggplot2")
library("hrbrthemes")
library ("dplyr")
```

## Exercise 1: Open data file

```{r data file}
somlit <- read.table("2022_SOMLIT.txt", header=TRUE)
somlit
```

-   What are the variables?

```{r data info}
variables <- names(somlit)
variables
```

-   How many observations are there in the database?

```{r number of observations}
observations <- dim(somlit)
observations
observations <- observations[1]
print(paste("number of observations is", observations))
```

## Exercise 2: time series analysis of chlorophyll-a concentration CHLA

-   Compute the basic descriptive statistics of CHLA

```{r descriptive stat}
attach(somlit) 
mean(CHLA)
median(CHLA)
sd(CHLA)
```

-   Generate the histogram of the relative frequency of CHLA (see Fig1) \*\* The relative frequency is the count over the total number of count

```{r histogram Base R / Lubac's solution}
hist(CHLA)
#this is not the histogram we want. needs to be relative frequency

Hist <- hist (CHLA, plot =FALSE)
Hist #see the data within the histogram. it shows the count and breaks(x-axis) that we need to compute relative frequency

# relative frequencies of CHLA is the count divided by the sum of the counts, this is the y value
(barplot(Hist$counts/sum(Hist$counts), 
  ylab="Relative frequency",
  names.arg=Hist$mids, xlab=expression(paste("CHLA(",mu,"g/L)")),
  main="Histogram of the relative frequency of CHLA"))
```

```{r ggplot solution}
ggplot(data = somlit, aes(x = CHLA)) +
  geom_histogram(aes(y = ..count../sum(..count..)), 
                 breaks =seq(0, 9, by = 1), fill = "gray", color = "black") +
  labs(x = expression(paste("CHLA (", mu, "g/L)")),
       y = "Relative frequency",
       title = "Histogram of the relative frequency of CHLA") +
  theme_minimal()
```

-   Generate the time series of CHLA (see Fig2)

```{r time series Base R / Lubac's solution}
#fix the date and time data type and formatting
somlit$DATE <- as.Date(somlit$DATE, format = "%d/%m/%Y")
#now the Date is no longer a character

plot(somlit$DATE, somlit$CHLA, type = "b",
     ylab = expression(paste("CHLA (", mu, "g/L)")),
     xlab = "Date",
     main = "Time Series of CHLA",
     xaxt = "n")  # Suppress the default x-axis

#to have the same x labels as Lubac's solution
# Customize the x-axis with specific breaks
axis.Date(1, at = seq(min(date), max(date), by = "2 years"), format = "%Y")
```

```{r ggplot solution}
# Generate the ggplot time series
ggplot(somlit, aes(x = DATE, y = CHLA)) +
  geom_line() + # Adds lines connecting the points
  geom_point() + # Adds points on the graph
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") + # Customizes the x-axis to show years at 2-year intervals
  labs(x = "Date", y = expression(paste("CHLA (", mu, "g/L)")), 
       title = "Time Series of CHLA") + 
  theme_minimal() + # Apply a minimal theme for clean styling
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
```

## Exercise 3: Analysis of seasonal variability of CHLA

-   Compute the mean and the standard deviation of CHLA per season (Spring, Summer, Autumn, Winter)

-   Display the basic descriptive statistics of CHLA per season

The **tapply()** function in R is used to apply a function (like mean, sum, or sd) to subsets of a vector, grouped by a factor or list of factors. It's a powerful tool for performing group-wise operations on data, similar to a "split-apply-combine" operation.

```{r Base R / Lubac's solution}
tapply(CHLA,Season,mean) # Calculate the mean CHLA value for each season
tapply(CHLA,Season,sd) # Calculate the standard deviation of CHLA for each season

#visualize the mean and sd using a boxplot
boxplot(CHLA~Season,ylab = expression(paste("CHLA (",mu,"g/L)")))

```

```{r ggplot solution}
#added some color for better visualization of the mean
ggplot(somlit, aes(x = Season, y=CHLA)) +
  geom_boxplot(fill = "lightblue")+
  stat_summary(fun = mean, geom = "point", shape = 20, size = 4, color = "darkblue") +  # Add mean as a dark blue point
  labs(x ="Season", y= expression(paste("CHLA (", mu, "g/L"))) +
  theme_minimal()  # Apply minimal theme for a clean plot

```

### Exercise 4: Detection of spring anomalies

-   Standardize CHLA variable (subtract the mean and divide by the standard deviation)
-   Find the date associated with CHLA values higher than 2.

```{r Base R / Lubac's solution}
#we are only interested in spring
CHLA.spring<-somlit[Season=="spring","CHLA"]
Date.spring<-somlit[Season=="spring","DATE"]

#formula for standardizing a variable is (variable - mean)/sd
CHLA.stand<-(CHLA.spring-mean(CHLA.spring))/sd(CHLA.spring)

#plot the standardized
boxplot(CHLA.stand)

#these are the dates where CHLA > 2
Date.anomalies <-Date.spring[CHLA.stand>2]
Date.anomalies
```

```{r dplry solution}

# Filter the dataset to keep only Spring data and display DATE and CHLA columns
spring_data <- somlit %>%
  filter(Season == "spring") %>%  # Filter for Spring season
  select(DATE, CHLA)  # Select only DATE and CHLA columns
spring_data

# Standardize the CHLA variable
mean_chla <- mean(spring_data$CHLA) # Compute the mean of CHLA
sd_chla <- sd(spring_data$CHLA)  # Compute the standard deviation of CHLA

# Create a new column for standardized CHLA
spring_data <- spring_data %>%
  mutate(CHLA_standardized = (spring_data$CHLA - mean_chla) / sd_chla)

# Plot the standardized CHLA
ggplot(spring_data, aes(y=CHLA)) +
  geom_boxplot()+
  labs(y= "Standardized CHLA") +
  theme_minimal() 

# Find dates associated with standardized CHLA > 2
anomalies <- spring_data %>%
  filter(CHLA_standardized > 2) %>%
  select(DATE, CHLA, CHLA_standardized)

# Display the results
anomalies

```

## Exercise 5: Relationship between CHLA and SPM

-   Display the relationship between CHLA and SPM (see the website : <https://www.r-graph-gallery.com/>)

```{r Lubac's solution using ggplot}

ggplot(somlit, aes(x=CHLA, y=SPM, color=Season))+
 	geom_point(size=4) +
    	theme_ipsum()
```

## Exercise 6: Reproduce the same plots

### dbinom() function

```         
-   plot(): Plots the binomial probabilities (fd) for n = 5.
-   fd \~ c(0:5): Y-values are fd (probabilities), and X-values are the number of successes (0 to 5).
-   xlim = c(0, 20): Sets the limits for the x-axis (from 0 to 20).
-   ylim = c(0, 0.40): Sets the limits for the y-axis (from 0 to 0.40).
-   pch = 15: Sets the plotting symbol (square) for the points.
-   col = coul[1]: Sets the color for the points (black).
-   cex = 2: Increases the size of the points.
-   lwd = 3: Increases the line width (applies to the points' border).
-   cex.axis = 2: Increases the size of axis tick labels.
```

This code calculates and plots the binomial distributions for different sample sizes (n = 5, n = 10, and n = 20), all with a success probability of p = 0.5.

```{r binomial distribution / Lubac's solution}
# Define a vector of colors for the different plots
coul <- c("black", "red", "green", "blue")

# Plot the binomial distribution for n = 5, p = 0.5
fd <- dbinom(c(0:5), 5, 0.5)
plot(fd ~ c(0:5), xlim = c(0, 20), ylim = c(0, 0.40), type = "p", ylab = " ", xlab = " ", pch = 15, cex = 2, lwd = 3, col = coul[1], cex.axis = 2)

# Add the binomial distribution for n = 10, p = 0.5
fd <- dbinom(c(0:10), 10, 0.5)
points(fd ~ c(0:10), type = "p", pch = 16, cex = 2, lwd = 3, col = coul[2])

# Add the binomial distribution for n = 20, p = 0.5
fd <- dbinom(c(0:20), 20, 0.5)
points(fd ~ c(0:20), type = "p", pch = 17, cex = 2, lwd = 3, col = coul[3])

# Create the legend text
legtxt <- c(expression(paste(italic(n), " = 5", sep = "")),
            expression(paste(italic(n), " = 10", sep = "")),
            expression(paste(italic(n), " = 20", sep = "")))

# Add the legend to the plot
legend("topright", legtxt, title = expression(paste(italic(p), " = 0.5", sep = "")),
       pch = c(15, 16, 17), col = c(coul[1], coul[2], coul[3]), cex = 2, bg = "white", inset = 0.050) #inset adds some space between the legend and the plot's edge

```

**Other notes**

-   type = "p": p stands for points. When you set type = "p", R will plot points at the specified (x, y) coordinates.

    -   type = "l": lines. This plots a line connecting the data points.
    -   type = "b": both. This plots both points and lines connecting them.
    -   type = "o": overplotted. This is similar to "b", but the points are plotted over the lines.
    -   type = "h": histogram-like. This creates vertical lines from the x-axis to the points (often used in bar plots).
    -   type = "n": no plotting. This sets up the axes and plotting area but doesn't plot anything. Useful when you want to add elements manually later, like using points() or lines().

### dnorm() and dt() function

```         
-   curve(dnorm, ...): The curve() function is used to plot a normal distribution curve (standard normal, N(0,1))

-   dnorm: The probability density function (PDF) of the standard normal distribution.

-   ylab = "" and xlab = "": No labels for the y-axis or x-axis.

-   lty = 5: Specifies the line type (dashed line) for this curve.

-   lwd = 3: Specifies the line width (thicker line).

-   add = FALSE: This creates a new plot (since it's the first curve).

-   col = coul[1]: Uses the first color (black) from the coul vector.

-   cex.axis = 2: Increases the size of the axis text.
```

This code plots the standard normal distribution (N(0,1)) and Student's t-distributions with degrees of freedom (n = 1, n = 2, n = 5). The curve() function is used to plot the probability density functions (PDFs) of these distributions.

```{r Lubac's solution}
# Plot the normal distribution curve (N(0,1))
curve(dnorm,from=-4,to=4,ylab="",xlab="",lty=5,lwd=3,add=F,col=coul[1],cex.axis=2)

# Add the Student's t-distribution curve with 1 degree of freedom
# dt(x, 1): The t-distribution with 1 degree of freedom.
# lty = 1: Solid line type for this curve.
curve(dt(x,1),from=-4,to=4,ylab="",xlab="",lty=1,lwd=3,add=T,col=coul[2])

# Student's t-distribution curve with 2 degrees of freedom
# lty = 2: Dashed line for this curve.
curve(dt(x,2),from=-4,to=4,ylab="",xlab="",lty=2,lwd=3,add=T,col=coul[3])

# Student's t-distribution curve with 5 degrees of freedom
# lty = 4: Dash-dot line type for this curve.
curve(dt(x,5),from=-4,to=4,ylab="",xlab="",lty=4,lwd=3,add=T,col=coul[4])

# Create the legend text with labels for different degrees of freedom and the normal distribution
{legtxt <- c(expression(paste(italic(n)," = 1", sep="")),
     expression(paste(italic(n)," = 2",sep="")),
     expression(paste(italic(n)," = 5",sep="")),
     expression(paste(italic(N),"(0,1)",sep="")))}

# Add the legend to the plot
{legend("topleft",legtxt,lty=c(1,2,4,5),lwd=3,col=c(coul[2],coul[3],coul[4],coul[1]), cex=1.6,bg="white",inset=0.0225)}
```

## Exercise 7

Packets of seeds of Zostera are commercially available. In a packet there are 15 seeds and each seed gives at most a foot of Zostera. Seagrasses have a probability of germination equal to 0.85.

**6.1.** What is the minimum number of packets that a PhD student must purchase to obtain at least 200 Zostera foot with a probability greater than or equal to 0.95?

**6.2.** The Phd student has 100 euros to purchase the seeds. A packet is sold 4.50 euros. How many feet should he expect to get? \*

**Given:**

p=0.85

P(X\>=200) =0.95

P(X\<200)=1-0.95=0.05

F(X=x)=P(X\<x)=0.05

The function **qbinom** provides the number of successes associated with a given probability

-   Example : qbinom(0.95,200,0.85) = 178 means in a case of a binomial distribution defined by a number of trials (n) equals to 200 and a probability of success for each trial (p) equals to 0.85, there is 95% of chance to have 178 successes. In this exercise, we have to determine the number of trials (n). Because the value of p is closed to 1, we can expected a value of n close to the number of successes (200). Thus, we can iterate n from 200 to 250 and checked the result provided by the function qbinom

```{r exercise 7}

## Plot the binomial quantile function for different numbers of seeds (n = 200 to 250)
plot(200:250,qbinom(0.05,200:250,0.85)) #x-axis is associated with n and y-axis with the 

# number of successes
abline(h=200) ## Add a horizontal line at 200 germinated seeds
abline(v=246) ## # Add a vertical line to show n = 246 seeds


fd <- dbinom(c(0:330),330,0.85)
# Plot the probability density function (PDF) for 330 seeds
plot(fd~c(0:330),xlim=c(240,300),ylim=c(0,.08),type="p",ylab=" ",xlab=" ",pch=15,cex=1,lwd=3,col=coul[1],cex.axis=2)
# This figure provides the PDF (p(X=x)). For example, there is 6.1% of chance to have 282 feet.

# Plot the cumulative distribution function (CDF) for 330 seeds
plot(cumsum(fd)~c(0:330),xlim=c(240,300),ylim=c(0,1),type="p",ylab=" ",xlab=" ",pch=15,cex=1,lwd=3,col=coul[1],cex.axis=2)
# This figure provides the cumulative distribution function (F(X=x)=P(X<=x)). For example, there is 55.4% of chance that the student obtain less than 282 feet of Zostera.

```

**Answer:**

-   To obtain at least 200 successes with a probability of 95%, the number of trials (n) has to be equal to 246.
-   The student has to purchase 17 (=246/15) packets of seeds to obtain at least 200 Zostera feet with a probability greater that or equal to 0.95.
-   with 100€, the student can purchase 330 (=100/4.5\*15) seeds
